---
title: "BacDive - Batch Download of traits"
author: "Konrad Herbst"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up

### Libraries
```{r}
library(tidyverse)
library(BacDive) # installed using `install.packages("BacDive", repos="http://R-Forge.R-project.org")` as noted on https://r-forge.r-project.org/R/?group_id=1573
library(pheatmap)
```

### Strain data
```{r}
strains_url <- "https://bacdive.dsmz.de/advsearch/csv"
strains_data_file <- str_c("../Data/dsmz-strains-", Sys.Date(), ".csv")
#strains_data_file <- str_c("../Data/dsmz-strains-", "2022-11-30", ".csv")
if(!(file.exists(strains_data_file))){
  download.file(strains_url, strains_data_file)
  strains <- read_csv(strains_data_file, skip = 2) %>%
    write_csv(file = strains_data_file)
}
strains <- read_csv(strains_data_file)
```

### BacDive connection
```{r}
# the file `.bacdive_api` contains the credentials and has as content:
### credentials <- c("bacdive user@email.com", "bacdive-user-passwort")
source(".bacdive_api") 
bacdive <- open_bacdive(credentials[[1L]], credentials[[2L]])
```

## Retrieve Phenotypes
A function to retrieve for a given set of strains (defined by their respective BacDive ID) `taxonomy`, `enzyme`, `metabolite utilization` and `metabolite production` information.
```{r}
retrieve_phenotypes <- function(bacdive, ids){
  id_fetched <- fetch(bacdive, ids)
  
  idr <- id_fetched$results
  
  res_annotated <- map_lgl(idr, ~length(.x[["Physiology and metabolism"]])>0)

  idr <- idr[res_annotated]
  if(length(idr)<1) return(NULL)
  
  res_phylo <- lapply(idr, function(x){
    id <- x[["General"]][["BacDive-ID"]]
    xx <- x[["Name and taxonomic classification"]]
    if(!is.null(xx)){
      ret <- tibble(
          id = id,
          domain = xx[["domain"]],
          phylum = xx[["phylum"]],
          class = xx[["class"]],
          order = xx[["order"]],
          family = xx[["family"]],
          genus = xx[["genus"]],
          species = xx[["species"]],
          strain = xx[["strain designation"]]
        )
      return(ret)
      }
    }) %>% bind_rows()
  
  #"oxygen tolerance"
  # res_ox <- tibblify(idr, tspec_df(
  #   id = tib_int(c("General", "BacDive-ID")),
  #   oxygen_tolerance = tib_chr(c("Physiology and metabolism", "oxygen tolerance", "oxygen tolerance"), required = FALSE)
  # )) %>% count(oxygen_tolerance)
  
  # "enzymes"
  res_E <- lapply(idr, function(x){
    id <- x[["General"]][["BacDive-ID"]]
    xx <- x[["Physiology and metabolism"]][["enzymes"]]
    if(!is.null(xx)){
      if(!is.null(names(xx))){
        ret <- tibble(
          id = id,
          enzyme_name = xx[["value"]],
          enzyme_ec = xx[["ec"]],
          enzyme_activity = if_else(is.null(xx[["activity"]]), NA_character_, xx[["activity"]])
        )
      } else {
        ret <- lapply(xx, function(y){
          tibble(
            id = id,
            
          enzyme_name = y[["value"]],
          enzyme_ec = y[["ec"]],
          enzyme_activity = if_else(is.null(y[["activity"]]), NA_character_, y[["activity"]])
          )
        }) %>% bind_rows()
      }
      ret
    }
  }) %>% bind_rows()
  if(nrow(res_E)>0){
    res_E <- res_E %>%
      mutate(feature = str_glue("E:{enzyme_name} ({enzyme_ec})"),
           value = recode(enzyme_activity, "+" = 2L, "-" = 0L, "+/-" = 1L)) %>%
      select(id, feature, value)
  }
  
  # "metabolite utilization"
  res_MU <- lapply(idr, function(x){
    id <- x[["General"]][["BacDive-ID"]]
    xx <- x[["Physiology and metabolism"]][["metabolite utilization"]]
    if(!is.null(xx)){
      if(!is.null(names(xx))){
        ret <- tibble(
          id = id,
          metab_util_name = xx[["metabolite"]],
          metab_util_kind = xx[["kind of utilization tested"]],
          metab_util_activity = xx[["utilization activity"]]
        )
      } else {
        ret <- lapply(xx, function(y){
          tibble(
            id = id,
          metab_util_name = y[["metabolite"]],
          metab_util_kind = y[["kind of utilization tested"]],
          metab_util_activity = y[["utilization activity"]]
          )
        }) %>% bind_rows()
      }
      ret
    }
  }) %>% bind_rows()
  if(nrow(res_MU)>0){
    res_MU <- res_MU %>%
    mutate(feature = str_glue("MU:{metab_util_name}_{metab_util_kind}"),
           value = recode(metab_util_activity, "+" = 2L, "-" = 0L, "+/-" = 1L)) %>%
      select(id, feature, value)
  }
  
  # "metabolite production"
  res_MP <- lapply(idr, function(x){
    id <- x[["General"]][["BacDive-ID"]]
    xx <- x[["Physiology and metabolism"]][["metabolite production"]]
    if(!is.null(xx)){
      if(!is.null(names(xx))){
        ret <- tibble(
          id = id,
          metab_prod_name = xx[["metabolite"]],
          metab_prod_production = xx[["production"]]
        )
      } else {
        ret <- lapply(xx, function(y){
          tibble(
            id = id,
            metab_prod_name = y[["metabolite"]],
            metab_prod_production = y[["production"]]
          )
        }) %>% bind_rows()
      }
      ret
    }
  }) %>% bind_rows()
  if(nrow(res_MP)>0){
    res_MP <- res_MP %>%
    mutate(feature = str_glue("MP:{metab_prod_name}"),
           value = recode(metab_prod_production, "yes" = 2L, "no" = 0L)) %>%
      select(id, feature, value)
  }
  
  #"Culture and growth conditions" -> temp?
  # "Physiology and metabolism" -> Halophilie?
  
  ## put everything together and label multiple entries as ambiguous
  pheno_x <- bind_rows(
      res_E,
      res_MU,
      res_MP
    )
  if(nrow(pheno_x)>0){
    pheno_x <- pheno_x %>%
    unique() %>%
    group_by(id, feature) %>%
    summarise(value = ifelse(n()>1, 1L, value))
    
    pheno_tab_long <- left_join(
      res_phylo,
      pheno_x,
      by = "id"
    ) %>% filter(!is.na(feature))
  
    return(pheno_tab_long)
  } else {
    return(NULL)
  }
}
```

Run phenotype retrieval for all strains.
```{r}
ids_all <- strains %>% pull(ID) %>% unique() %>% sort()

ids_alls <- split(ids_all, ceiling(seq_along(ids_all)/99))[1:3]
# ids_alls <- ids_alls[sample(1:length(ids_alls), 80)]
length(ids_alls)

phenotype_data_file <- str_c("../Data/bacdive-pheno_mat-", Sys.Date(), ".RData")
#phenotype_data_file <- str_c("../Data/bacdive-pheno_mat-", "2022-11-30", ".RData")
if(!(file.exists(phenotype_data_file))){
  ids_alls_res <- lapply(seq_along(ids_alls), function(i){
    print(str_glue("{i} / {length(ids_alls)}"))
    print(ids_alls[[i]])
    res <- retrieve_phenotypes(bacdive, ids_alls[[i]])
    Sys.sleep(0.2)
    return(res)
  }) %>%
    bind_rows()
  save(ids_alls_res, file = phenotype_data_file)
}
```

## Exploration
```{r}
load("../Data/bacdive-pheno_mat-2022-11-30.RData")
```

Top features
```{r}
ids_alls_res %>%
  count(feature) %>%
  arrange(desc(n)) %>% View()
```

Top species
```{r}
ids_alls_res %>%
  count(id) %>%
  arrange(desc(n)) %>% View()
```

```{r}
#ggplot(ids_alls_res) +
#  geom_tile(aes(y=as.factor(id), x= feature, fill = as.factor(value)))
pheno_mat <- ids_alls_res %>%
  pivot_wider(names_from = feature, values_from = value)
dim(pheno_mat)
pheno_mat_noNA <- pheno_mat %>% select(-c(2:9)) %>% column_to_rownames("id")
pheno_mat_noNA[is.na(pheno_mat_noNA)] <- -1L
```

Percantage of empty cells
```{r}
pheno_mat_noNA_tab <- table(t(pheno_mat_noNA))
pheno_mat_noNA_tab[["-1"]]/sum(pheno_mat_noNA_tab)*100
sum(pheno_mat_noNA_tab) - pheno_mat_noNA_tab[["-1"]]
```

```{r}
features <- ids_alls_res %>%
  separate(feature, c("feature_class", "feature_desc"), ":", remove = FALSE) %>%
  count(feature, feature_class, feature_desc) %>%
  arrange(desc(n))# %>%
  #write_tsv("features.tsv")
```

## Feature selection
```{r}
pheno_mat_filt <- pheno_mat_noNA[, colSums(pheno_mat_noNA>-1)>=400]
pheno_mat_filt <- pheno_mat_filt[rowSums(pheno_mat_filt>-1)>=30, ]
dim(pheno_mat_filt)
pheatmap(pheno_mat_filt, color = c(rep("#222222", 49), colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(100)), 
main = "Filtered phenotypearray (all species with at least 30 features which are present in at least 400 species)", filename = "../Results/heatmap-pheno_mat_filt.png", width = 20, height = 13)
#pheatmap(pheno_mat_filt, kmeans_k = 1000)
```

```{r}
pheno_mat_select <- ids_alls_res %>%
  filter(str_detect(feature, "builds acid from")) %>%
  pivot_wider(names_from = feature, values_from = value) %>%
  select(-c(2:9)) %>% column_to_rownames("id")
pheno_mat_select <- pheno_mat_select[rowSums(!is.na(pheno_mat_select))>=30, ]
pheno_mat_select <- pheno_mat_select[, colSums(!is.na(pheno_mat_select))>=400]
dim(pheno_mat_select)

# remove all species/features with at least one NA value
pheno_mat_select <- na.omit(pheno_mat_select)
# remove all features, which have no variability in information
pheno_mat_select <- pheno_mat_select[,apply(pheno_mat_select, 2, sd)>0]
# remove all samples, which have no variability in information
pheno_mat_select <- pheno_mat_select[apply(pheno_mat_select, 1, sd)>0,]
#pheno_mat_select[is.na(pheno_mat_select)] <- -1L
dim(pheno_mat_select)

pheatmap(pheno_mat_select,
         main = "Selected phenotypes")#,
         #filename = "heatmap-pheno_mat_filt.png", width = 20, height = 13)

correlating_features <- which(cor(pheno_mat_select, method = "spearman")>0.74, arr.ind = TRUE)

correlating_features[correlating_features[,"row"] != correlating_features[,"col"],]

pheno_mat_select_red <- pheno_mat_select[,-c(8,33,41)]
cor(pheno_mat_select_red, method = "spearman") %>% as.numeric() %>% hist
dim(pheno_mat_select_red)
pheatmap(pheno_mat_select_red,
         main = "Selected phenotypes",
         filename = "heatmap-pheno_mat_select_red.png", width = 10, height = 8)
# pheno_mat_select_red %>%
#   rownames_to_column("id") %>%
#   write_tsv("20221205-pheno_mat_select_red-builds_acid_from.tsv")
```

Using doubly lexcial sorting to identify potential subset.
```{r}
source("doubly_lexical_ordering.R")
Mpheno <- ids_alls_res %>%
  filter(str_detect(feature, "builds acid from")) %>%
  pivot_wider(names_from = feature, values_from = value) %>%
  select(-c(2:9)) %>% column_to_rownames("id")
Mpheno <- Mpheno[rowSums(!is.na(Mpheno))>=30, ]
Mpheno <- Mpheno[, colSums(!is.na(Mpheno))>=400]
dim(Mpheno)
dlo <- dlordering(binarize_matrix(Mpheno))

pheatmap(Mpheno[dlo$R, dlo$C], cluster_rows = FALSE, cluster_cols = FALSE, na_col = "black",
         main = "'Builds acid from'-phenotypes (all species with at least 30 features which are present in at least 400 species)",
         filename = "heatmap-pheno_mat_select_red-builds_acid_from-dlo.png", width = 10, height = 8)

Mpheno <- ids_alls_res %>%
  pivot_wider(names_from = feature, values_from = value) %>%
  select(-c(2:9)) %>% column_to_rownames("id")
Mpheno <- Mpheno[rowSums(!is.na(Mpheno))>=50, ]
Mpheno <- Mpheno[, colSums(!is.na(Mpheno))>=600]
dim(Mpheno)
dlo <- dlordering(binarize_matrix(Mpheno))

pheatmap(Mpheno, cluster_rows = FALSE, cluster_cols = FALSE, na_col = "black")
pheatmap(Mpheno[dlo$R, dlo$C], cluster_rows = FALSE, cluster_cols = FALSE, na_col = "black", main = "")

```

<2022-12-12> After discussing with Taylor, we decided to restrict ourself to a submatrix which would omit ambiguous values as the current algorithm would be needed to be adapted. 
```{r}
Mpheno <- ids_alls_res %>%
  filter(str_detect(feature, "builds acid from")) %>%
  pivot_wider(names_from = feature, values_from = value) %>%
  select(-c(2:9)) %>% column_to_rownames("id")
Mpheno <- Mpheno[rowSums(!is.na(Mpheno))>=30, ]
Mpheno <- Mpheno[, colSums(!is.na(Mpheno))>=400]
Mpheno[Mpheno == 1] <- NA_integer_ 
dim(Mpheno)
dlo <- dlordering(binarize_matrix(Mpheno))

pheatmap(Mpheno[dlo$R, dlo$C], cluster_rows = FALSE, cluster_cols = FALSE, na_col = "black",
         main = "'Builds acid from'-phenotypes (all species with at least 30 features which are present in at least 400 species)")#,
         #filename = "heatmap-pheno_mat_select_red-builds_acid_from-dlo.png", width = 10, height = 8)

pheno_mat_select <- ids_alls_res %>%
  filter(str_detect(feature, "builds acid from"),
         ## manually remove features with high number of ambiguous values
         !(feature %in% c("MU:sorbitol_builds acid from",
                          "MU:methyl beta-D-glucopyranoside_builds acid from",
                          "MU:pullulan_builds acid from",
                          "MU:alpha-cyclodextrin_builds acid from"))) %>%
  pivot_wider(names_from = feature, values_from = value) %>%
  select(-c(2:9)) %>% column_to_rownames("id")
pheno_mat_select <- pheno_mat_select[rowSums(!is.na(pheno_mat_select))>=30, ]
pheno_mat_select <- pheno_mat_select[, colSums(!is.na(pheno_mat_select))>=400]
pheno_mat_select[pheno_mat_select == 1] <- NA_integer_ 
dim(pheno_mat_select)
# remove all features with at least one NA value
pheno_mat_select <- pheno_mat_select[!(rowSums(is.na(pheno_mat_select))>=1),]
dim(pheno_mat_select)

# remove all features, which have no variability in information
pheno_mat_select <- pheno_mat_select[,apply(pheno_mat_select, 2, sd)>0]
# remove all samples, which have no variability in information
pheno_mat_select <- pheno_mat_select[apply(pheno_mat_select, 1, sd)>0,]
#pheno_mat_select[is.na(pheno_mat_select)] <- -1L
dim(pheno_mat_select)

pheatmap(pheno_mat_select,
         main = "Selected phenotypes")#,
         #filename = "heatmap-pheno_mat_filt.png", width = 20, height = 13)

correlating_features <- which(cor(pheno_mat_select, method = "spearman")>0.74, arr.ind = TRUE)

correlating_features[correlating_features[,"row"] != correlating_features[,"col"],]

pheno_mat_select_red <- pheno_mat_select[,-c(35,43,44)]
cor(pheno_mat_select_red, method = "spearman") %>% as.numeric() %>% hist
dim(pheno_mat_select_red)
pheatmap(pheno_mat_select_red,
         main = "Selected phenotypes",
         filename = "../Results/20221212-heatmap-pheno_mat_select_red.png", width = 10, height = 8)
pheno_mat_select_red %>%
  rownames_to_column("id") %>%
  write_tsv("../Data/20221212-pheno_mat_select_red-builds_acid_from.tsv")
```

## Session info
```{r}
sessionInfo()
```
