---
title: "MASS Figures"
author: "Konrad Herbst (with code from Elena Forchielli)"
date: "3/1/2021; revised 11/2023"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Set-Up and Helper Functions
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)
library(patchwork)
library(ggpubr)
```

```{r}
# compute Shannon entropy
entropy <- function(target) {
  freq <- table(target)/length(target)
  # vectorize
  vec <- as.data.frame(freq)[,2]
  #drop 0 to avoid NaN resulting from log2
  vec<-vec[vec>0]
  #compute entropy
  -sum(vec * log2(vec))
}
```

```{r}
my_breaks <- function(x, b = 5){
  x <- as.numeric(x)
  ret <- sapply(x, function(xx){
    ifelse(xx %% b == 0 | xx == x[[1]] | xx == x[[length(x)]], as.character(xx), '')
  })
  return(ret)
}
```

```{r}
# read balanced accuracy scores (calculated over all RFs per predictor)
read_performance_scores <- function(fname){
  ret <- read_csv(fname) %>%
    dplyr::rename(num_predictors = `# of predictors`)
  col <- colnames(ret)
  col <- col[[length(col)]]
  ret <- ret %>%
    group_by(num_predictors, i, predictors) %>%
    summarise("mean_{col}" := mean( .data[[col]]  )) %>%
    ungroup() %>%
    mutate(mass_prediction = case_when(
      i == 0 ~ "MASS",
      i == 1 ~ "entropy",
      TRUE ~ "random"
    )) %>%
    arrange(num_predictors)
  return(ret)
}
```


# DATASET 1 (marine)
## Data Loading
```{r}
# read MASS result
marine_mass_p <- read_csv('../Results/MIP_classify_z_all_11.csv') %>%
  rownames_to_column("num_responses") %>%
  mutate(num_responses = as.numeric(num_responses),
    num_predictors = max(num_responses)-num_responses+1) %>%
  select(-num_responses) %>%
  pivot_longer(-num_predictors, names_to = 'response', values_to = c('pred')) %>%
  group_by(response) %>% mutate(x = sum(pred)) %>% ungroup() %>%
  arrange(desc(x)) %>% mutate(response = fct_inorder(response))

# binary growth profiles
gp <- readxl::read_xlsx('../Data/growth_profiles_binary.xlsx')
```

## Figure: Marine dataset main figure
### Phenotype matrix

```{r}
gp_ent <- lapply(gp[,-1], entropy) 
ent <- data.frame(medium = names(gp_ent),
                  entropy = unlist(gp_ent))
```

```{r}
### Figure raw phenotype matrix as heatmap ###

# order rows and columns with hclust
marine_hm_clust_org <- hclust(dist(column_to_rownames(gp, "strain"), method = "euclidian"), method = "complete")
marine_hm_clust_cond <- hclust(dist(t(select(gp, -strain)), method = "euclidian"), method = "complete")

marine_hm <- gp %>%
  pivot_longer(-strain, names_to = "condition", values_to = "phenotype")

fig_marine_hm <- marine_hm %>%
  mutate(strain = fct_relevel(strain, marine_hm_clust_org$labels[marine_hm_clust_org$order]),
    condition = fct_relevel(condition, marine_hm_clust_cond$labels[marine_hm_clust_cond$order])) %>%
  ggplot(., aes(x = condition, y = strain)) +
  geom_tile(aes(fill = as.factor(phenotype)), colour = "gray") +
  scale_fill_viridis(discrete = TRUE, labels = c("0" = "Negative", "1" = "Positive"), name = "Growth") +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = "bottom",
        axis.title.x = element_blank(), axis.title.y = element_blank())
fig_marine_hm
```

```{r}
marine_hm_cond_var <- marine_hm %>%
  group_by(condition) %>%
  summarise(var = var(phenotype)) %>%
  ungroup() %>%
  arrange(desc(var)) %>%
  mutate(rank = fct_reorder(condition, var))
```

### MASS applied
```{r warning=FALSE}
### Figure heatmap selected predictors
# ordered by predictor selection prevalence
fig_marine_mass_p <- marine_mass_p %>%
  mutate(num_predictors = as.factor(num_predictors)) %>%
  ggplot(., aes(x = num_predictors, y = response)) +
  geom_tile(aes(fill = as.factor(pred)), colour = "gray") +
  labs(x = "Number of Predictors") +
  scale_fill_manual(values = c("0" = "black", "1" = "gray"), guide = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title.y = element_blank(), axis.text.y = element_blank())
fig_marine_mass_p + theme_minimal()

### Figure bar graphs: Entropy
fig_marine_mass_ent <- ent %>%
  mutate(medium = factor(medium, levels = levels((marine_mass_p)$response))) %>%
  ggplot(aes(x = medium, y = entropy, fill = medium)) +
  geom_col(width = 1) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.y = element_text(hjust = 0.5, colour = "black")) +
  scale_fill_viridis(discrete = TRUE, direction = -1) +
  scale_y_continuous(breaks = c(0,.5,1)) +
  # scale_y_continuous(breaks = seq(0,1,.2)) +
  guides(fill = "none") +
  labs(x = "",
       y = "Shannon Entropy") +
  coord_flip()
fig_marine_mass_ent
fig_marine_mass_p + fig_marine_mass_ent
```

### RF performance
```{r}
marine_baseline_stats <- read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-acc.csv") %>%
  left_join(., read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-bacc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-f1weighted.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-jaccard.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-cohen_kappa.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_marine_baseline_20230913_stats-mcc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
    mutate(num_predictors = factor(num_predictors, levels = 1:10))

# just MCC
fig_marine_mass_mcc <- marine_baseline_stats %>%
  ggplot(., aes(num_predictors, mean_mcc, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(drop = FALSE) +#, breaks = my_breaks) +
  labs(#title = "Yeast dataset: RF performance ~ #predictors",
       x = "# predictors",
       y = "mean(MCC over\nall RF predictions)") +
  theme_classic() +
  theme(legend.position = "bottom")
fig_marine_mass_mcc
```

### Plot assembly
```{r}
png(filename = "../Results/fig-marine-mass.png", width = 4000, height = 3500, res = 300)
wrap_plots(A = fig_marine_hm,
           B = fig_marine_mass_p,
           C = fig_marine_mass_ent, 
           D = fig_marine_mass_mcc,
           design = layout <- '
ABC
ADD
', heights = c(1,1), widths = c(3,3,1)) +
  plot_annotation(tag_levels = "A") & 
  theme(plot.tag = element_text(size = 18))
dev.off()
```

## Supplementary Figure: all RF performance scores
```{r}
png(filename = "../Results/fig-supp-marine-rf_perf_all.png", width = 3500, height = 2000, res = 300)
fig_marine_rf_perf_all <- marine_baseline_stats %>%
  pivot_longer(starts_with("mean_"), names_to = "performance_score", values_to = "performance_value") %>%
  ggplot(., aes(num_predictors, performance_value, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(drop = FALSE) +
  labs(title = "Marine dataset: RF performance ~ #predictors",
       x = "# predictors",
       # y = "mean(balanced accuracy over all RF predictions)",
       y = "mean(respect. performance score over all RF predictions)") +
    facet_wrap(facets = vars(performance_score), ncol = 2) +
  theme_classic() +
  theme(legend.position = "bottom")
fig_marine_rf_perf_all
dev.off()
```

## Supplementary Figure: Per-Response plot
```{r}
### Figure dot plots: MCC
marine_mcc_js <- read_csv("../Results/MIP_marine_baseline_20230913_stats-mcc_js.csv") %>%
    dplyr::rename(num_predictors = `# of predictors`) %>%
  group_by(num_predictors, i, predictors, response) %>%
  summarize(mean_mcc = mean(mcc), sd_mcc = sd(mcc)) %>%
  ungroup() %>%
  mutate(num_predictors = if_else(num_predictors==11, "C", as.character(num_predictors))) %>% 
  mutate(num_predictors = factor(num_predictors, levels = c(as.character(1:10), "C"))) %>%
  mutate(response = factor(response, levels = levels(fct_reorder(marine_mass_p$response, marine_mass_p$x))))

fig_marine_mass_perattr_mcc <- marine_mcc_js %>%
  ggplot(aes(x = as.factor(num_predictors), y = mean_mcc, color = response)) +
  geom_point(aes(size = sd_mcc), alpha = .75) +
  theme_classic() +
  theme(text = element_text(size = 14),
        legend.justification=c(1,0), 
        legend.position=c(1,0)) +
  scale_color_viridis(discrete = TRUE) +
  facet_wrap(~response) +
  guides(color = "none") +
  labs(y = "MCC",
       x = "Number of Predictors",
       color = "Medium",
       size = "Standard Deviation")
png(filename = "../Results/fig-supp-marine-rf_perf_MCC_perResponse.png", width = 3500, height = 2500, res = 300)
fig_marine_mass_perattr_mcc
dev.off()
```

# DATASET 2 (bacdive1)
## Data Loading
```{r}
bacdive1dat <- read_tsv("../Data/20221212-pheno_mat_select_red-builds_acid_from.tsv") %>%
  pivot_longer(-id, names_to = "response") %>%
  mutate(response = str_match(response, "MU:(.+)_builds acid from")[,2]) %>%
  pivot_wider(names_from = "response")

# read in f1 scores
bacdive1f1 <- read_csv("../Data/MIP_classify_RF_f1_score_agg-bacdive1.csv") %>%
  dplyr::rename(num_predictors = `# of predictors`) %>%
  mutate(response = str_match(response, "MU:(.+)_builds acid from")[,2]) %>%
  select(-std) %>% dplyr::rename(f1_score = mean) %>%
  arrange(num_predictors)
```

## Figure: BacDive1 dataset main figure
### Phenotype matrix
```{r}
### Figure raw phenotype matrix as heatmap ###

# order rows and columns with hclust
bacdive1_hm_clust_org <- hclust(dist(column_to_rownames(bacdive1dat, "id"), method = "euclidian"), method = "complete")
bacdive1_hm_clust_cond <- hclust(dist(t(select(bacdive1dat, -id)), method = "euclidian"), method = "complete")
bacdive1_hm <- bacdive1dat %>%
  pivot_longer(-id, names_to = "condition", values_to = "phenotype") %>%
  mutate(id = fct_relevel(as.character(id), bacdive1_hm_clust_org$labels[bacdive1_hm_clust_org$order]),
    condition = fct_relevel(condition, bacdive1_hm_clust_cond$labels[bacdive1_hm_clust_cond$order])
    )

fig_bacdive1_hm <- bacdive1_hm %>%
  ggplot(., aes(x = condition, y = id)) +
  geom_tile(aes(fill = as.factor(phenotype))) +
  scale_fill_viridis_d(labels = c("0" = "Negative", "2" = "Positive"), name = "Fermentation") +
  labs(y = "Species") +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        legend.text = element_text(size = 10, color = "black"),
        legend.title = element_text(size = 10, color = "black"),
        legend.key.size = unit(0.8, "lines"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.line.y = element_line(colour = "black", linewidth = rel(1)))
fig_bacdive1_hm
```


```{r}
### Figure heatmap marginal conditions ###
# percentage of bacdive species fermenting on each carbon source
fig_bacdive1_hm_cond <- bacdive1_hm %>%
  ggplot(aes(x = condition, fill = as.factor(phenotype))) +
  geom_bar(position = "fill", width = 1) +
  scale_y_continuous(labels = scales::percent, breaks = c(0,.5,1)) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title = element_blank(), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_text(hjust = 0.5, color = "black")) +
    #theme(text = element_blank()) +
  scale_fill_viridis_d(guide = NULL) +
  guides(fill = "none") +
  labs(x = "Carbon Source",
       y = "Percentage of Species")
fig_bacdive1_hm_cond
```

```{r}
### Figure heatmap marginal organisms ###
# the order of the bars (strains) matches the order of the strains in the heat map
fig_bacdive1_hm_org <- bacdive1_hm %>%
  ggplot(aes(x = id, fill = as.factor(phenotype))) +
  geom_bar(position = "fill", width = 1) +
  scale_y_continuous(labels = scales::percent, breaks = c(0,.5,1)) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_viridis_d(guide = NULL) +
  #guides(fill = "none") +
  labs(x = "Species",
       y = "Carbon Sources")+
  coord_flip()
fig_bacdive1_hm_org
```

```{r}
# png(filename = "fig-bacdive1-heatmap.png", width = 4500, height = 3000, res = 300)
 wrap_plots(A = fig_bacdive1_hm_cond,
           B = guide_area(), 
           C = fig_bacdive1_hm,
           D = fig_bacdive1_hm_org,
           design = layout <- '
AB
CD
', heights = c(1, 9), widths = c(6,1), guides = 'collect') +
  plot_annotation(tag_levels = "A") & 
  theme(plot.tag = element_text(size = 18))
# dev.off()
```

### MASS applied
```{r}
# format data
bacdive1f1_wide <- bacdive1f1 %>%
  pivot_wider(names_from = response, values_from = f1_score)
# c sources selected as predictors will have NA, change to 0
bacdive1f1_wide[is.na(bacdive1f1_wide)] <- 1
# anything w f1 score was a response & all f1 < 1, so convert all < 1 to 0, meaning NOT chosen as predictor
bacdive1f1_wide[bacdive1f1_wide < 1] <- 0
```

```{r}
### Figure fig_bacdive1_mass ###
# format for heatmap
bacdive1_mass_p <- bacdive1f1 %>%
  complete(num_predictors, response) %>%
  mutate(pred = if_else(is.na(f1_score), 1, 0)) %>%
  group_by(response) %>% mutate(x = sum(pred),
                                xfirst = suppressWarnings( min(num_predictors[pred==1]) )
                                ) %>% ungroup() %>%
  arrange(desc(x)) %>% mutate(response = fct_inorder(response))

# ordered by predictor selection prevalence
fig_bacdive1_mass_p <- bacdive1_mass_p %>%
  mutate(num_predictors = as.factor(num_predictors)) %>%
  ggplot(., aes(x = num_predictors, y = response)) +
  geom_tile(aes(fill = as.factor(pred)), colour = "gray") +
  labs(x = "Number of Predictors") +
  scale_x_discrete(breaks = my_breaks) +
  scale_y_discrete(position = "right") +
  scale_fill_manual(values = c("0" = "black", "1" = "gray"), guide = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title.y = element_blank(), 
        # axis.text.y = element_blank()
        )
fig_bacdive1_mass_p

# subselection on the first 8 p's
fig_bacdive1_mass_p8 <- bacdive1_mass_p %>%
  filter(num_predictors<9) %>%
  group_by(response) %>%
  filter(sum(pred)>0) %>%
  mutate(y = min(num_predictors[pred>0])) %>% ungroup() %>%
  mutate(response = fct_reorder(response, y, .desc = TRUE)) %>%
  mutate(num_predictors = as.factor(num_predictors)) %>%
  ggplot(., aes(x = num_predictors, y = response)) +
  geom_tile(aes(fill = as.factor(pred)), colour = "gray") +
  labs(x = "Number of Predictors") +
  scale_fill_manual(values = c("0" = "black", "1" = "gray"), guide = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title.y = element_blank(), 
        # axis.text.y = element_blank()
        )
fig_bacdive1_mass_p8


# ordered by first p selected as predictor
fig_bacdive1_mass_p_xfirst <- bacdive1_mass_p %>%
  arrange(xfirst) %>% mutate(response = fct_inorder(response)) %>%
  mutate(num_predictors = as.factor(num_predictors)) %>%
  ggplot(., aes(x = num_predictors, y = response)) +
  geom_tile(aes(fill = as.factor(pred)), colour = "gray") +
  labs(x = "Number of Predictors") +
  scale_x_discrete(breaks = my_breaks) +
  scale_y_discrete(position = "right") +
  scale_fill_manual(values = c("0" = "black", "1" = "gray"), guide = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title.y = element_blank(), axis.text.y = element_blank())
fig_bacdive1_mass_p_xfirst +
  (ent %>%
     mutate(carbon_source = factor(carbon_source, levels = levels((bacdive1_mass_p%>% mutate(response =fct_reorder(response, xfirst)))$response))) %>%
  ggplot(aes(x = carbon_source, y = entropy, fill = carbon_source)) +
  geom_col(width = 1) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.y = element_text(hjust = 0.5, colour = "black")) +
  scale_fill_viridis(discrete = TRUE, direction = -1) +
  scale_y_continuous(breaks = c(0,.5,1)) +
  # scale_y_continuous(breaks = seq(0,1,.2)) +
  guides(fill = "none") +
  labs(x = "",
       y = "Shannon Entropy") +
  coord_flip())
```

```{r}
# compute entropy
e <- lapply(select(bacdive1dat, -id), entropy) 
# format data
# match order to heatmap of c sources selected as predictors
ent <- tibble(carbon_source = names(e),
                  entropy = unlist(e)) %>%
  mutate(carbon_source = factor(carbon_source, levels(bacdive1_mass_p$response)))
```

```{r}
### Figure bar graphs: Entropy ###
fig_bacdive1_mass_ent <- ent %>%
  ggplot(aes(x = carbon_source, y = as.numeric(entropy), fill = carbon_source)) +
  geom_col(width = 1) +
  scale_fill_viridis_d(direction = -1) +
  scale_y_continuous(breaks = c(0,.5,1)) +
  guides(fill = "none") +
  labs(x = "",
       y = "Shannon Entropy") +
  coord_flip() +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.y = element_blank()
        # axis.text.y = element_text(hjust = 0.5, colour = "black")
        )
fig_bacdive1_mass_ent

fig_bacdive1_mass_p + fig_bacdive1_mass_ent
```

### RF Performance
```{r}
bacdive1_baseline_stats <- read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-acc.csv") %>%
  left_join(., read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-bacc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-f1weighted.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-jaccard.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-cohen_kappa.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_bacdive1_baseline_20230523_stats-mcc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
    mutate(num_predictors = factor(num_predictors, levels = 1:45))

# just MCC
fig_bacdive1_mass_mcc <- bacdive1_baseline_stats %>%
  ggplot(., aes(num_predictors, mean_mcc, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n<=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(drop = FALSE, breaks = my_breaks) +
  labs(#title = "Bacdive1 dataset: RF performance ~ #predictors",
       x = "# predictors",
       y = "mean(MCC over\nall RF predictions)") +
  theme_classic() +
  theme(legend.position = "bottom")
fig_bacdive1_mass_mcc
```


### Plot assembly
```{r}
# png(filename = "fig-bacdive1-mass.png", width = 5500, height = 5500, res = 300)
pdf(file = "../Results/fig-bacdive1-mass.pdf", width = 20, height = 22)
wrap_plots(A = fig_bacdive1_hm_cond,
           B = guide_area(), 
           C = fig_bacdive1_hm,
           D = fig_bacdive1_hm_org,
           E = fig_bacdive1_mass_p,
           F = fig_bacdive1_mass_ent,
           G = (fig_bacdive1_mass_mcc + theme(axis.title.x = element_text(vjust = 35)) + plot_layout(guides = 'keep')),
           H = (fig_bacdive1_mass_p8 + scale_y_discrete(position = "right") + ggtitle(label = "Fermented Carbon Source")),
           I = grid::textGrob("Relevant Carbohydrate Monomers"),
           J = grid::textGrob("momomers influx metabolic map"),
           design = '
AAAABEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
HHIIIJJJJJ
HHIIIJJJJJ
HHIIIJJJJJ
', guides = 'collect') +
  plot_annotation(tag_levels = list(c("A", "", "", "B", "C", "D", "E", "F", "G"))) & 
  theme(plot.tag = element_text(size = 18))
dev.off()
```

## Supplementary Figure: all RF performance scores
```{r}
png(filename = "../Results/fig-supp-bacdive1-rf_perf_all.png", width = 3500, height = 2000, res = 300)
fig_bacdive1_rf_perf_all <- bacdive1_baseline_stats %>%
  pivot_longer(starts_with("mean_"), names_to = "performance_score", values_to = "performance_value") %>%
  ggplot(., aes(num_predictors, performance_value, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  # scale_color_discrete(name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(drop = FALSE, breaks = ~my_breaks(.x, 2)) +
  labs(title = "Bacdive1 dataset: RF performance ~ #predictors",
       x = "# predictors",
       # y = "mean(balanced accuracy over all RF predictions)",
       y = "mean(respective performance score over all RF predictions)") +
  facet_wrap(facets = vars(performance_score), ncol = 2) +
  theme_classic() +
  theme(legend.position = "bottom")
fig_bacdive1_rf_perf_all
dev.off()
```


# DATASET 3 (yeast)
## Data Loading
```{r}
### import and format yeast data ###
rawDat <- readxl::read_xlsx("../Data/summary_2ndEdition.xlsx", sheet = "Sheet1", trim_ws = TRUE) %>% 
  filter(!strain == "dummyVars") %>% # second row has the dummy vars, remove
  rename(species = strain)

# get list of carbon sources included in analysis
cSources <- readxl::read_xlsx("../Data/summary_2ndEdition.xlsx", sheet = "Sheet2", trim_ws = TRUE) 
# subset features
rawDat <- rawDat %>% select(c(cSources$carbon_source, "species"))
# remove the word "growth" from column names
rawDat <- rawDat %>%
  setNames(gsub(" growth", "", names(.)))

# replace characters with numbers, see above
yeastdat <- rawDat %>%
  mutate(across(everything(), ~replace(., . %in%  c("D", "V", "W") , 1))) %>%
  mutate(across(everything(), ~replace(., . == "+", 2))) %>%
  mutate(across(everything(), ~replace(., . == "-", 0))) %>%
  mutate(across(everything(), ~replace(., . == "?", NA)))

# remove rows with unknown feature values (NAs) and no variance
yeastdat <- yeastdat %>%
  drop_na()

# read MASS result
yeast_mass_p <- read_csv('../Results/MIP_classify_z_all76_greedy2.csv') %>%
  rownames_to_column("num_responses") %>%
  mutate(num_responses = as.numeric(num_responses),
    num_predictors = max(num_responses)-num_responses+1) %>%
  select(-num_responses) %>%
  pivot_longer(-num_predictors, names_sep = '_', names_to = c('response', 'x'), values_to = c('pred')) %>%
 # mutate(pred = if_else(resp==0, 1, 0)) %>%
  filter(x == '2.0') %>% select(-x) %>% # '_1.0' and '2.0' columns contain the same values
  filter(response != 'D-Glucose', num_predictors<35) %>% # filter out unused variables
  group_by(response) %>% mutate(x = sum(pred)) %>% ungroup() %>%
  arrange(desc(x)) %>% mutate(response = fct_inorder(response))
```


## Figure: Yeast dataset main figure
### Phenotype matrix

```{r}
### Figure raw phenotype matrix as heatmap ###

# order rows and columns with hclust
yeast_hm_clust_org <- hclust(dist(column_to_rownames(yeastdat, "species"), method = "euclidian"), method = "complete")
yeast_hm_clust_cond <- hclust(dist(t(select(yeastdat, -species)), method = "euclidian"), method = "complete")
yeast_hm <- yeastdat %>%
  pivot_longer(-species, names_to = "condition", values_to = "phenotype") %>%
  mutate(species = fct_relevel(as.character(species), yeast_hm_clust_org$labels[yeast_hm_clust_org$order]),
    condition = fct_relevel(condition, yeast_hm_clust_cond$labels[yeast_hm_clust_cond$order])
    )

fig_yeast_hm <- yeast_hm %>%
  ggplot(., aes(x = condition, y = species)) +
  geom_tile(aes(fill = as.factor(phenotype))) +
  scale_fill_viridis_d(labels = c("0" = "Negative", "1" = "Variable, weak,\n delayed", "2" = "Positive"), name = "Growth") +
  labs(y = "Species") +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        legend.text = element_text(size = 10, color = "black"),
        legend.title = element_text(size = 10, color = "black"),
        legend.key.size = unit(0.8, "lines"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.line.y = element_line(colour = "black", linewidth = rel(1)))
fig_yeast_hm
```


```{r}
### Figure heatmap marginal conditions ###
# percentage of bacdive species fermenting on each carbon source
fig_yeast_hm_cond <- yeast_hm %>%
  ggplot(aes(x = condition, fill = as.factor(phenotype))) +
  geom_bar(position = "fill", width = 1) +
  scale_y_continuous(labels = scales::percent, breaks = c(0,.5,1)) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title = element_blank(), 
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_text(hjust = 0.5, color = "black")) +
  scale_fill_viridis_d(guide = NULL) +
  guides(fill = "none") +
  labs(x = "Carbon Source",
       y = "Percentage of Species")
fig_yeast_hm_cond
```

```{r}
### Figure heatmap marginal organisms ###
# the order of the bars (strains) matches the order of the species in the heat map
fig_yeast_hm_org <- yeast_hm %>%
  ggplot(aes(x = species, fill = as.factor(phenotype))) +
  geom_bar(position = "fill", width = 1) +
  scale_y_continuous(labels = scales::percent, breaks = c(0,.5,1)) +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_fill_viridis_d(guide = NULL) +
  #guides(fill = "none") +
  labs(x = "Species",
       y = "Carbon Sources") +
  coord_flip()
fig_yeast_hm_org
```

```{r}
# png(filename = "fig-bacdive1-heatmap.png", width = 4500, height = 3000, res = 300)
 wrap_plots(A = fig_yeast_hm_cond,
           B = guide_area(), 
           C = fig_yeast_hm,
           D = fig_yeast_hm_org,
           design = layout <- '
AB
CD
', heights = c(1, 9), widths = c(6,1), guides = 'collect') +
  plot_annotation(tag_levels = "A") & 
  theme(plot.tag = element_text(size = 18))
# dev.off()
```

### MASS applied
```{r}
### Figure fig_yeast_mass ###
fig_yeast_mass_p <- yeast_mass_p %>%
  mutate(num_predictors = as.factor(num_predictors)) %>%
  ggplot(., aes(x = num_predictors, y = response)) +
  geom_tile(aes(fill = as.factor(pred)), colour = "gray") +
  labs(x = "Number of Predictors") +
  scale_x_discrete(breaks = my_breaks) +
  scale_y_discrete(position = "right") +
  scale_fill_manual(values = c("0" = "black", "1" = "gray"), guide = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 14, color = "black"),
        axis.title.y = element_blank(), 
        axis.text.y.right = element_text(hjust = .5)
        )
fig_yeast_mass_p
```


```{r}
# compute entropy
e <- lapply(select(yeastdat, -species), entropy) 
# format data
# match order to heatmap of c sources selected as predictors
ent <- tibble(carbon_source = names(e),
                  entropy = unlist(e)) %>%
  filter(carbon_source != "D-Glucose") %>%
  mutate(carbon_source = factor(carbon_source, levels(yeast_mass_p$response)))
```

```{r}
### Figure bar graphs: Entropy ###
fig_yeast_mass_ent <- ent %>%
  ggplot(aes(x = carbon_source, y = as.numeric(entropy), fill = carbon_source)) +
  geom_col(width = 1) +
  scale_fill_viridis_d(direction = -1) +
  scale_y_continuous(breaks = c(0,.5,1)) +
  guides(fill = "none") +
  labs(x = "",
       y = "Shannon Entropy") +
  coord_flip() +
  theme_classic() +
  theme(text = element_text(size = 14, color = "black"),
        # axis.text.y = element_text(hjust = 0.5, colour = "black"),
        axis.text.y = element_blank()
        )
fig_yeast_mass_ent
```

### RF performance
```{r}
yeast_baseline_stats <- read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-acc.csv") %>%
  left_join(., read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-bacc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-f1weighted.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-jaccard.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-cohen_kappa.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
  left_join(., read_performance_scores("../Results/MIP_yeast_baseline_20230411_stats-mcc.csv"), by = c("num_predictors", "i", "predictors", "mass_prediction")) %>%
    mutate(num_predictors = factor(num_predictors, levels = 1:36))

# just MCC
fig_yeast_mass_mcc <- yeast_baseline_stats %>%
  ggplot(., aes(num_predictors, mean_mcc, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(breaks = my_breaks) +
  labs(#title = "Yeast dataset: RF performance ~ #predictors",
       x = "# predictors",
       y = "mean(MCC over\nall RF predictions)") +
  theme_classic() +
  theme(legend.position = "bottom")
fig_yeast_mass_mcc
```

### Plot assembly
```{r}
# plot assembly
png(filename = "../Results/fig-yeast-mass.png", width = 5500, height = 4300, res = 300)
wrap_plots(A = fig_yeast_hm_cond,
           B = guide_area(), 
           C = fig_yeast_hm,
           D = fig_yeast_hm_org,
           E = fig_yeast_mass_p,
           F = fig_yeast_mass_ent,
           G = (fig_yeast_mass_mcc + theme(axis.title.x = element_text(vjust = 25)) + plot_layout(guides = 'keep')),
           design = '
AAAABEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDEEEEF
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
CCCCDGGGGG
', guides = 'collect') +
  plot_annotation(tag_levels = list(c("A", "", "", "B", "C", "D", "E"))) & 
  theme(plot.tag = element_text(size = 18))
dev.off()
```


## Supplementary Figure: all RF performance scores
```{r}
png(filename = "../Results/fig-supp-yeast-rf_perf_all.png", width = 3500, height = 2000, res = 300)
fig_yeast_rf_perf_all <- yeast_baseline_stats %>%
  pivot_longer(starts_with("mean_"), names_to = "performance_score", values_to = "performance_value") %>%
  ggplot(., aes(num_predictors, performance_value, color = mass_prediction)) +
  geom_point(aes(shape = mass_prediction, size = mass_prediction), position = position_jitter(width = .2, height = 0), alpha = .6) +
  stat_summary(aes(group = mass_prediction), geom = 'line', fun = 'mean', linewidth = 1) +
  scale_color_brewer(palette = "Set1", name = "RF with", labels = c("random" = "random predictors (n=300)", "MASS" = "MASS-selected predictors", "entropy" = "max-entropy-selected predictors")) +
  scale_shape_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_size_manual(values = c("random" = .4, "MASS" = 2, "entropy" = 2), guide = 'none') +
  # scale_alpha_manual(values = c("random" = 1, "MASS" = 16, "entropy" = 16), guide = 'none') +
  scale_x_discrete(drop = FALSE, breaks = my_breaks) +
  labs(title = "Yeast dataset: RF performance ~ #predictors",
       x = "# predictors",
       # y = "mean(balanced accuracy over all RF predictions)",
       y = "mean(respect. performance score over all RF predictions)") +
    facet_wrap(facets = vars(performance_score), ncol = 2) +
  theme_classic() +
  theme(legend.position = "bottom")
fig_yeast_rf_perf_all
dev.off()
```
# Supplementary Figure: Predictor selection frequency vs. Shannon Entropy (all DATASETs)
```{r}
# DATASET 1
marine_nsel <- marine_mass_p %>% group_by(response) %>% summarize(nsel = sum(pred))
marine_ent <- gp %>% pivot_longer(-strain, names_to = 'medium', values_to = 'growth') %>%
  group_by(medium) %>%
  summarize(entropy = entropy(growth))
# DATASET 2
bacdive1_nsel <- bacdive1_mass_p %>% group_by(response) %>% summarize(nsel = sum(pred))
bacdive1_ent <- bacdive1dat %>% pivot_longer(-id, names_to = 'medium', values_to = 'growth') %>%
  group_by(medium) %>%
  summarize(entropy = entropy(growth))
# DATASET 3
yeast_nsel <- yeast_mass_p %>% group_by(response) %>% summarize(nsel = sum(pred))
yeast_ent <- yeastdat %>% pivot_longer(-species, names_to = 'medium', values_to = 'growth') %>%
  group_by(medium) %>%
  summarize(entropy = entropy(growth))

# assembling everything together
all_nsel_ent <- bind_rows(
    left_join(marine_nsel, marine_ent, by = c('response' = 'medium')) %>% mutate(dataset = 'marine'),
    left_join(bacdive1_nsel, bacdive1_ent, by = c('response' = 'medium')) %>% mutate(dataset = 'bacdive1'),
    left_join(yeast_nsel, yeast_ent, by = c('response' = 'medium'))  %>% mutate(dataset = 'yeast'))

plot_nsel_ent <- function(ds){
  p <- ggplot(ds, aes(x = entropy, y = nsel)) +
  geom_point() +
  stat_cor(method = "pearson") +
  scale_y_continuous(breaks = scales::pretty_breaks(10)) +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  labs(y = "Times Selected as Predictor",
       x = "Shannon Entropy")
  return(p)
}

filter(all_nsel_ent, dataset == 'marine') %>% plot_nsel_ent() /
filter(all_nsel_ent, dataset == 'bacdive1') %>% plot_nsel_ent() 

# plot assembly
png(filename = "../Results/fig-supp-all-nsel_ent.png", width = 4000, height = 3500, res = 300)
wrap_plots(A = filter(all_nsel_ent, dataset == 'marine') %>% plot_nsel_ent(),
           B = filter(all_nsel_ent, dataset == 'bacdive1') %>% plot_nsel_ent(), 
           C = filter(all_nsel_ent, dataset == 'yeast') %>% plot_nsel_ent(),
           D = plot_spacer(),
           E = plot_spacer(),
           design = '
AABB
AABB
DCCE
DCCE
', guides = 'collect') +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 18))
dev.off()
```

# Session Info
```{r}
sessionInfo()
```